---
title: "EDA"
format: html
author: "Evan Whitfield"
---

# Water Potability Analysis

[Click here](https://www.kaggle.com/datasets/adityakadiwal/water-potability) for information about the data we will be analyzing. A summary has been given below.

## About Dataset
"Access to safe drinking-water is essential to health, a basic human right and a component of effective policy for health protection. This is important as a health and development issue at a national, regional and local level."

## Variables

 - pH value:
PH is an important parameter in evaluating the acid–base balance of water

 - Hardness:
Hardness is mainly caused by calcium and magnesium salts.

 - Solids (Total dissolved solids - TDS):
A wide range of inorganic and some organic minerals or salts such as potassium, calcium, sodium, bicarbonates, chlorides, magnesium, sulfates etc can be produced giving un-wanted taste and diluted color.

 - Chloramines:
Chlorine and chloramine are the major disinfectants used in public water systems

 - Sulfate:
Sulfates are naturally occurring substances that are found in minerals, soil, and rocks

 - Conductivity:
Pure water is not a good conductor of electric current but rather is a good insulator. 

 - Organic_carbon:
Total Organic Carbon (TOC) in source waters comes from decaying natural organic matter (NOM) as well as synthetic sources.

 - Trihalomethanes:
THMs are chemicals which may be found in water treated with chlorine.

 - Turbidity:
The turbidity of water depends on the quantity of solid matter present in the suspended state.

 - Potability:
Indicates if water is safe for human consumption

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(psych)
library(ggplot2)

water_data <- read_csv("Data/water_potability.csv") |>
  mutate(Potability = factor(Potability,
                             levels = c(0, 1),
                             labels = c("Not Potable", "Potable")))

water_data
```

```{r}
str(water_data)
```

```{r}
describe(water_data)
```


```{r}
colSums(is.na(water_data))
```

```{r}
water_data_trimmed <- water_data |>
  drop_na()

describe(water_data_trimmed)
```

```{r}
water_data_trimmed |>
  group_by(Potability) |>
  summarize(count = n())
```

```{r}
water_data_trimmed |>
  group_by(Potability) |>
  summarize(mean_ph = mean(ph), 
            median_ph = median(ph))
```


```{r}
water_data_trimmed |>
  group_by(Potability) |>
  summarize(across(where(is.numeric),
      .fns = list(
        mean = mean,
        median = median,
        var = var,
        sd = sd,
        IQR = IQR
      ),
      .names = "{.fn}_{.col}"
    )
  ) |>
  pivot_longer(
    -Potability,
    names_to = c("Variable", "Statistic"),
    names_sep = "_",
    values_to = "Value"
  ) |>
  pivot_wider(
    names_from = Potability,
    values_from = Value
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 3)))
```


```{r}
water_data_trimmed |>
  pivot_longer(
    cols = c(ph, Hardness, Solids, Chloramines, Sulfate,
             Conductivity, Organic_carbon,
             Trihalomethanes, Turbidity),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  ggplot(aes(x = Value, fill = Potability)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Variable, scales = "free") +
  labs(
    title = "Density Plots of Variables by Potability",
    x = "Value",
    y = "Density",
    fill = "Potability"
  )
```


```{r}
water_data_trimmed |>
  select(where(is.numeric)) |>
  cor(use = "complete.obs") |>
  round(3)
```

```{r}
water_data_trimmed |>
  pivot_longer(
    cols = c(Hardness, Solids, Chloramines, Sulfate,
             Conductivity, Organic_carbon,
             Trihalomethanes, Turbidity),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  ggplot(aes(x = ph, y = Value, color = Potability)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(
    title = "Relationship Between pH and Other Variables",
    x = "pH",
    y = "Value",
    color = "Potability"
  )
```



[Click here for the Modeling Page](Modeling.html)
