---
title: "EDA"
format: html
author: "Evan Whitfield"
---

# Water Potability Analysis

[Click here](https://www.kaggle.com/datasets/adityakadiwal/water-potability) for information about the data we will be analyzing. A summary has been given below.

## About Dataset
"Access to safe drinking-water is essential to health, a basic human right and a component of effective policy for health protection. This is important as a health and development issue at a national, regional and local level."

## Variables

 - **pH value:**
PH is an important parameter in evaluating the acid–base balance of water

 - **Hardness:**
Hardness is mainly caused by calcium and magnesium salts.

 - **Solids** (Total dissolved solids - TDS):
A wide range of inorganic and some organic minerals or salts such as potassium, calcium, sodium, bicarbonates, chlorides, magnesium, sulfates etc can be produced giving un-wanted taste and diluted color.

 - **Chloramines:**
Chlorine and chloramine are the major disinfectants used in public water systems

 - **Sulfate:**
Sulfates are naturally occurring substances that are found in minerals, soil, and rocks

 - **Conductivity:**
Pure water is not a good conductor of electric current but rather is a good insulator. 

 - **Organic_carbon:**
Total Organic Carbon (TOC) in source waters comes from decaying natural organic matter (NOM) as well as synthetic sources.

 - **Trihalomethanes:**
THMs are chemicals which may be found in water treated with chlorine.

 - **Turbidity:**
The turbidity of water depends on the quantity of solid matter present in the suspended state.

 - **Potability:**
Indicates if water is safe for human consumption

```{r}
#| include: false
library(readr)
library(tidyr)
library(dplyr)
library(psych)
library(ggplot2)
```

## Read in Data
First we will read in the data. Since we know that Potability is our only categorical variable - and only has values of 0 and 1 - we will make it a factor variable. This will help us by giving us the text on if the water is potable or not instead of having to remember what a 0 or a 1 meant.
```{r}
water_data_raw <- read_csv("Data/water_potability.csv") |>
  mutate(Potability = factor(Potability,
                             levels = c(0, 1),
                             labels = c("Not Potable", "Potable")))

water_data_raw
```
Here we can see that all of our numeric variables are read in as `dbl` type accurately. There are 3276 observations, however, we can see that there are NA values in the pH and Sulfate column.

## Beginning Data Dive
```{r}
str(water_data_raw)
```
The structure of our data looks solid. Now let us look at the individual variables to see if everything looks good there.
```{r}
describe(water_data_raw)
```
The only noteworthy pieces of info at this point is that `n` is different for some of the variables due to the presence of NA entries. Also, not that it will mean much to us, but due to changing Potability to a factor variable, the values are now counted as 1s and 2s instead of 0s and 1s. This is why our mean is allowed to be greater than 1, despite the variable only being coded as 0s and 1s originally. Since Potability is considered a categorical variable, we do not need these summary statistics of it, but at first glance it appears that the data would have been entered or read incorrectly. All is fine!

```{r}
colSums(is.na(water_data_raw))
```
Three variables have NA entries: pH, Sulfate, and Trihalomethanes. Here, we made a decision to remove all observations with NA values for any variable. We could have kept all observations, but removed these variables from our investigation. However, pH is known to be a very important value for liquids and should be considered for Potability.
```{r}
water_data <- water_data_raw |>
  drop_na()

describe(water_data)
```
After dropping all NA observations, we still have over 2000 data observations. This should be more than plenty. If it is found that three variables with NA entries are not important to our models, we can add back in the removed data and recalculate the models, if need be.

## Data Summary

Potability will be used as our response variable for our model fitting. We will group by Potability status for our summary.
```{r}
water_data |>
  group_by(Potability) |>
  summarize(count = n())
```
Out of our 2011 observations, 1200 were Not Potable water and 811 were Potable.

### Uni-Variate Exploration

```{r}
water_data |>
  group_by(Potability) |>
  summarize(mean_ph = mean(ph), 
            median_ph = median(ph))
```
It appears that Potable water had a slightly higher mean and median pH than Not Potable water. However, the values are fairly close with standard deviations that might lead us to believe that they are not statistically significant. That is not our purpose in this study, but it could warrant further study.

#### Numerical Summaries

```{r}
water_data |>
  group_by(Potability) |>
  summarize(across(where(is.numeric),
      .fns = list(
        mean = mean,
        median = median,
        var = var,
        sd = sd,
        IQR = IQR
      ),
      .names = "{.fn}_{.col}")
  ) |>
  pivot_longer(
    -Potability,
    names_to = c("Variable", "Statistic"),
    names_sep = "_",
    values_to = "Value"
  ) |>
  pivot_wider(
    names_from = Potability,
    values_from = Value
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 3)))
```
Now that we have the summary information about each variable, we can see that there is not much difference between the Potable and Not Potable water. A further, deeper T-test could tell us if any of these values are statistically significant (again, not our purpose). However, on quick glance, there does seem to be any obviously significant differences for us to focus on.

#### Graphical Summaries

```{r}
water_data |>
  pivot_longer(
    cols = c(ph, Hardness, Solids, Chloramines, Sulfate,
             Conductivity, Organic_carbon,
             Trihalomethanes, Turbidity),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  ggplot(aes(x = Value, fill = Potability)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Variable, scales = "free") +
  labs(
    title = "Density Plots of Variables by Potability",
    x = "Value",
    y = "Density",
    fill = "Potability"
  )
```
This is backed up by the graphs above. Each density plot is grouped by Potability status. Many of the variables have a similar distribution for each Potability status. Sulfate, pH, and hardness seem to be the three variables with the most notable differences. However, the distributions have similar center values, but different spreads. Initial intuition would have guessed that the centers might also be different leading to a change in Potability status. That does not appear to be the case.

### Bi-Variate Exploration

```{r}
water_data |>
  select(where(is.numeric)) |>
  cor(use = "complete.obs") |>
  round(3)
```
Since there is not any obvious signs from a signal variable to influence Potability status, maybe there is a connection between two variables. The above `correlation matrix` shows that there is not a strong correlation between any variables. The strongest magnitude correlation is between Sulfate and Solids at r = -0.163. This is still a weak correlation, and most of the other r values are less than 0.02 in magnitude.

```{r}
water_data |>
  pivot_longer(
    cols = c(Hardness, Solids, Chloramines, Sulfate,
             Conductivity, Organic_carbon,
             Trihalomethanes, Turbidity),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  ggplot(aes(x = ph, y = Value, color = Potability)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(
    title = "Relationship Between pH and Other Variables",
    x = "pH",
    y = "Value",
    color = "Potability"
  )
```

These graphs show the relationship between pH and all other numeric variables. The strongest correlation between these is pH vs. Hardness at r = 0.103. Every scatter plot shows mostly a random scatter of points.

Because the other variables have similar correlation values, it stands to reason that the scatter plots of other variables on the x-axis would yield similar graphs.

There does not appear to be a single or pair of variables that would accurately predict the potability of water. We will need to create a model that incorporates many (if not all) of these variables to predict Potability.

[Click here for the Modeling Page](Modeling.html)
